Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-06-07 TEX/draft4.tex
latest updates

We compute the average autocorrelation of the micrographs (equivalently, the average of their power spectra). This is a particularly simple computation. In the methods section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the power spectrum of the unknown image itself. Mathematically, it is easy to show that the quality of this estimate improves steadily as the amount of data grows, regardless of noise level. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{bauschke2004rrr,elser2017rrr}.\footnote{\TODO{Quid of symmetries?}} RRR is initialized with a picture of the physicist Isaac Newton and iterates to produce the estimate.\footnote{\TODO{Initializing with a face is debatable: it might help the algorithm. On the other hand, I agree that it supports the notion that we don't get trapped too easily.}}

@@

\TODO{Do we want to note here that EM used to be perceived as computationally out of reach back when it was proposed [cite Fred], yet is now the de facto standard method as exemplified by RELION? Do we also want to stress that models have been refined over decades?}

@@

For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generate $P$ micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we place Einstein's image of size $50\times 50$  in random locations, while preserving the separation condition~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation constraint and locations picked so far.
On average, $700$ images are placed in each micrograph.   
Then, i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ is added, inducing an $\SNR$ of approximately $1/20$.
An example of a micrograph's excerpt is presented in the right panel of Figure~\ref{fig:micro_example}.
%Different micrographs are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

In this experiment, we assume we know the noise level $\sigma$ as well as the total number of occurrences of the target image across all micrographs.
In start contrast with the 1-D setup, the second-order autocorrelation determines almost any target image uniquely, up to reflection through the origin~\cite{hayes1982reconstruction,bendory2017fourier}.\footnote{\TODO{Should we cite \cite{bendory2017fourier} here? I'm not against citing it somewhere, but for the specific claim being justified here, it seems the older reference should be enough.}} This is because the second-order autocorrelations correspond to the Fourier magnitudes of the signal through the 2-D Fourier transform. 
Therefore, we estimate the signal's Fourier magnitudes (or power spectrum) from the Fourier magnitudes of the micrographs, at the cost of one 2-D fast Fourier transform (FFT) per micrograph. These can be computed highly efficiently and in parallel.

To recover the target image from the estimated power spectrum, we use a phase retrieval algorithm called relaxed-reflect-reflect (RRR). This algorithm iterates the map
\begin{align*}
	z & \leftarrow z + \beta (P_2(2P_1(z) - z) - P_1(z))
\end{align*}
on an image $z$ of size $2L\times 2L$.
We set the parameter $\beta$ to 1.
The map is designed so that, if the estimated power spectrum is exact, then fixed points contain Einstein's image in the upper-left corner of size $L \times L$, and zeros elsewhere. The operator $P_2(z)$ combines the Fourier phases of  the current estimation $z$ with the estimated Fourier magnitudes. The operator $P_1(z)$ zeros out all entries of $z$ outside the $L\times L$ upper-left corner.

>>>>

We compute the average autocorrelation of the micrographs (equivalently, the average of their power spectra). This is a particularly simple computation. In the methods section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the power spectrum of the unknown image itself. Mathematically, it is easy to show that the quality of this estimate improves steadily as the amount of data grows, regardless of noise level. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{bauschke2004rrr,elser2017rrr}. RRR is initialized with a picture of the physicist Isaac Newton and iterates to produce the estimate, up to a reflection ambiguity.\footnote{\TODO{Initializing with a face is debatable: it might help the algorithm. On the other hand, I agree that it supports the notion that we don't get trapped too easily.}}

@@

\TODO{Do we want to note here that EM used to be perceived as computationally out of reach back when it was proposed [cite Fred], yet is now the de facto standard method as exemplified by RELION? Do we also want to stress that models have been refined over decades? Both points are rather defensive in nature.	}

@@

For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generate $P$ micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we place Einstein's image (of zero mean) of size $50\times 50$  in random locations, while preserving the separation condition~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation constraint and locations picked so far.
On average, $700$ images are placed in each micrograph.   
Then, i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ is added, inducing an $\SNR$ of approximately $1/20$.
An example of a micrograph's excerpt is presented in the right panel of Figure~\ref{fig:micro_example}.
%Different micrographs are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

In this experiment, we assume we know the noise level $\sigma$ and the total number of occurrences of the target image across all micrographs.
In start contrast with the 1-D setup, the second-order autocorrelation determines almost any target image uniquely, up to reflection through the origin~\cite{hayes1982reconstruction,bendory2017fourier}.\footnote{\TODO{Should we cite \cite{bendory2017fourier} here? I'm not against citing it somewhere, but for the specific claim being supported here, it seems the older reference should be enough.}} This is because the second-order autocorrelations correspond to the Fourier magnitudes of the signal through the 2-D Fourier transform. 
Therefore, we estimate the signal's Fourier magnitudes (or power spectrum) from the Fourier magnitudes of the micrographs, at the cost of one 2-D fast Fourier transform (FFT) per micrograph. These can be computed highly efficiently and in parallel.

To recover the target image from the estimated power spectrum, we use a phase retrieval algorithm called relaxed-reflect-reflect (RRR). This algorithm iterates the map
\begin{align*}
	z & \leftarrow z + \beta (P_2(2P_1(z) - z) - P_1(z))
\end{align*}
on an image $z$ of size $2L\times 2L$.
We set the parameter $\beta$ to 1.
The map is designed so that, if the estimated power spectrum is exact, then fixed points contain Einstein's image in the upper-left corner of size $L \times L$, \TODO{possibly reflected}, and zeros elsewhere. The operator $P_2(z)$ combines the Fourier phases of  the current estimation $z$ with the estimated Fourier magnitudes. The operator $P_1(z)$ zeros out all entries of $z$ outside the $L\times L$ upper-left corner.
